Challenges to the Two-Infall Scenario by Large Stellar Age Catalogs

### 2025-07-01 src/tex/ms.tex

Subsequent studies refined the two-infall scenario to reproduce abundance data across the disk \citep[e.g.,][]{chiappini_abundance_2001,chiappini_oxygen_2003}. Others have explored the SN Ia delay-time distribution \citep{matteucci_effect_2009,palicio_analytic_2023}, galactic fountains \citep{spitoni_effects_2009}, radial gas flows \citep{spitoni_effects_2011,palla_chemical_2020}, variations in the star formation efficiency \citep{spitoni_effects_2011,palla_chemical_2020}, radial stellar migration \citep{spitoni_effect_2015,palla_mgfe_2022}, azimuthal abundance variations \citep{spitoni_2d_2019}, and pre-enriched gas infall \citep{palla_chemical_2020,spitoni_remind_2024}. \citet{spitoni_beyond_2023} and \citet{palla_mapping_2024} proposed a third gas accretion event in the last $\sim3$ Gyr to match the inferred star formation history from {\it Gaia} \citep{ruiz-lara_recurrent_2020} and to explain the recent abundance evolution (or lack thereof) in the Solar neighborhood. Most recently, \citet{hegedus_reconstructing_2025} constrained the infall timescales using chemical abundances for $\sim400,000$ stars from the 19th data release of the Sloan Digital Sky Survey (SDSS; \todo{Hawkins et al. in prep}).

@@

Our second age catalog utilizes the [C/N]--age relation calibrated by \todo{Roberts et al.\ (in prep)} for red giant branch (RGB) and red clump stars. The relationship relies on the mass-dependent level of mixing during first dredge-up \citep[FDU;][]{iben_stellar_1967} to map the correlation of stellar mass, and hence age, with surface chemistry. This method has the benefit of providing age estimates for luminous giants ($\log g<2.5$), which increases the sample size at larger distances from the Sun. Limitations from the efficiency of FDU mixing and the RGB age--mass relationship mean the ages are not trustworthy outside the range $1\sim10\Gyr$, but this is the age range most useful for our purposes anyway. Additional mixing effects in low-metallicity stars %later in their RGB evolution 
also prevent the relation from being applied to luminous giant and red clump stars with $\mathFeH<-0.4$; stars on the lower RGB do not suffer from this problem. The median propagated uncertainty for the [C/N]-derived ages is $\sim1\Gyr$; however, as noted by \todo{Roberts et al.\ (in prep)}, the propagated errors underestimate the true age dispersion, so we enhance the uncertainties by 40\% (see Table \ref{tab:uncertainties}). With this relationship, we estimate ages for \num{113464} stars across the disk, including \num{20995} in the Solar neighborhood.

@@

% \begin{figure*}
%     \centering
%     \includegraphics{figures/spitoni_comparison.pdf}
%     \caption{Chemical abundance tracks and distributions from one-zone models which use our fiducial parameters (solid curves) and the MCMC-derived parameters from \citet[][dashed curves]{spitoni_apogee_2021} at three different galactocentric radii. All models use the $y/Z_\odot=1$ yield set and corresponding outflow prescription (see Table \ref{tab:yields}). The time markers and marginal panels are the same as in Figure \ref{fig:twoinfall-parameters}. We also include a 2-D heatmap of the APOGEE stellar abundances in each panel for reference. APOGEE stars are restricted to (a) $2\leq R_{\rm gal}<6$ kpc, (b) $6\leq R_{\rm gal}<10$ kpc, and (c) $10\leq R_{\rm gal}<14$ kpc, and $0\leq |z|<2$ kpc in each panel.}
%     \label{fig:spitoni-comparison}
% \end{figure*}

% \todo{Highlight differences between our models and those of \citet{spitoni_apogee_2021} in Figure \ref{fig:spitoni-comparison}.}

@@

\subsection{Dilution \& Re-enrichment}
\label{sec:age-abundance}\begin{figure}
    \centering
    \includegraphics[width=\onecolumn]{figures/yield_outflow.pdf}
    \caption{The abundance evolution of three one-zone models with different yield sets and outflow mass-loading factors. Table \ref{tab:yields} presents the population-averaged yields for each model. The gray points plot the abundances of APOGEE stars with NN ages from \citet{leung_variational_2023} from the Solar neighborhood ($7\leq R_{\rm gal}<9\kpc$, $0\leq|z|<0.5\kpc$). The black points with error bars indicate the mode of the abundance data in 1 Gyr-wide age bins, and the gray error bars along the bottom of each panel indicate the median age and abundance errors as a function of age.}
    \label{fig:yield-outflow}
\end{figure}

@@

We further highlight the discrepant age patterns in the bottom panels of Figure \ref{fig:lmr-ages}, which compare the overall stellar age distribution against that of the locally metal rich (LMR) stars, defined here as $\mathFeH\ge+0.1$.\footnote{
    While the amplitude of the peaks is sensitive to the precise location of the LMR cutoff, the dearth of intermediate-age stars is robust to adjustments of $\pm0.05\dex$.
} For APOGEE, the distributions are similar, both peaking near $\sim5\Gyr$, although very few of the LMR stars have ages $\gtrsim10\Gyr$. Our two-infall models predict an overall age distribution that is similar to the data, but both models predict a distinctly bimodal age distribution for LMR stars. The two populations represent the metal-rich endpoints of the two successive infall epochs. The trough between the modes lies at $\sim5\Gyr$ for both models, right where the APOGEE distribution peaks. It is difficult to predict a unimodal age distribution with the available parameter space. A dearth of intermediate-age, high-metallicity stars is a natural outcome of a substantial dilution event, which is a defining feature of the two-infall scenario. None of the adjustments that we explored in Figure \ref{fig:abundance-evolution-params} substantially increase the proportion of intermediate-age stars.

@@

% Radial gas flows allow GCE models to produce a radial metallicity gradient in the absence of mass-loaded outflows.
Radial gas flows introduce multiple variables, such as the speed, direction, and angular momentum of the flow, which remain relatively unconstrained in the MW. \citet{spitoni_effects_2011} studied the effects of radial gas flows in the two-infall scenario and found that an inward radial flow with velocity $v_R\sim0-4\kms$ can steepen the radial abundance gradient to match observations. Subsequent studies \citep{palla_chemical_2020,palla_mapping_2024} implemented a $\sim-1\kms$ radial gas flow in order to reproduce the abundance gradient in the absence of Galactic outflows. Confirming these results, \todo{Johnson et al. (in prep)} find that an inward gas flow with a flat velocity profile has a similar effect on the radial metallicity gradient as the exponential outflow prescription in our models. We believe that switching from outflows to radial gas flows would not solve any of our models' issues regarding the age--abundance relation, [O/Fe] distributions, or stellar age distributions.

% Some two-infall studies \citep[e.g.,][]{spitoni_effects_2011,palla_chemical_2020,palla_mapping_2024} implement inward radial gas flows with velocity $\sim1$ km s$^{-1}$ in order to reproduce the radial abundance gradient without Galactic outflows. \citet{spitoni_effects_2011} find that a two-infall model of the disk without gas exchange produces a radial metallicity gradient which is too shallow. They implement an inward radial gas flow on the order of $\sim0-4$ km s$^{-1}$ which varies with radius, and find that it improves agreement with the observed gradient. However, they found that a variable star formation efficiency with radius in combination with a gas density threshold for star formation could also reproduce the observed gradient without radial flows.

>>>>

Subsequent studies refined the two-infall scenario to reproduce abundance data across the disk \citep[e.g.,][]{chiappini_abundance_2001,chiappini_oxygen_2003}. Others have explored the SN Ia delay-time distribution \citep{matteucci_effect_2009,palicio_analytic_2023}, galactic fountains \citep{spitoni_effects_2009}, radial gas flows \citep{spitoni_effects_2011,palla_chemical_2020}, variations in the star formation efficiency \citep{spitoni_effects_2011,palla_chemical_2020}, radial stellar migration \citep{spitoni_effect_2015,palla_mgfe_2022}, azimuthal abundance variations \citep{spitoni_2d_2019}, and pre-enriched gas infall \citep{palla_chemical_2020,spitoni_remind_2024}. \citet{spitoni_beyond_2023} and \citet{palla_mapping_2024} proposed a third gas accretion event in the last $\sim3$ Gyr to match the inferred star formation history from {\it Gaia} \citep{ruiz-lara_recurrent_2020} and to explain the recent abundance evolution (or lack thereof) in the Solar neighborhood. Most recently, \citet{hegedus_reconstructing_2025} constrained the infall timescales using chemical abundances for $\sim400,000$ stars from the 19th data release of the Sloan Digital Sky Survey (SDSS; \todo{Hawkins et al. 2025, in preparation}).

@@

Our second age catalog utilizes the [C/N]--age relation calibrated by \todo{Roberts et al.\ (2025, in preparation)} for red giant branch (RGB) and red clump stars. The relationship relies on the mass-dependent level of mixing during first dredge-up \citep[FDU;][]{iben_stellar_1967} to map the correlation of stellar mass, and hence age, with surface chemistry. This method has the benefit of providing age estimates for luminous giants ($\log g<2.5$), which increases the sample size at larger distances from the Sun. Limitations from the efficiency of FDU mixing and the RGB age--mass relationship mean the ages are not trustworthy outside the range $1\sim10\Gyr$, but this is the age range most useful for our purposes anyway. Additional mixing effects in low-metallicity stars %later in their RGB evolution 
also prevent the relation from being applied to luminous giant and red clump stars with $\mathFeH<-0.4$; stars on the lower RGB do not suffer from this problem. The median propagated uncertainty for the [C/N]-derived ages is $\sim1\Gyr$; however, as noted by \todo{Roberts et al.\ (2025, in preparation)}, the propagated errors underestimate the true age dispersion, so we enhance the uncertainties by 40\% (see Table \ref{tab:uncertainties}). With this relationship, we estimate ages for \num{113464} stars across the disk, including \num{20995} in the Solar neighborhood.

@@

@@

\subsection{Dilution \& Re-enrichment}
\label{sec:age-abundance}

\begin{figure}
    \centering
    \includegraphics[width=\onecolumn]{figures/gas_abundance_evolution.pdf}
    \caption{The abundance evolution of three one-zone models with different yield sets and outflow mass-loading factors. Table \ref{tab:yields} presents the population-averaged yields for each model. The gray points plot the abundances of APOGEE stars with NN ages from \citet{leung_variational_2023} from the Solar neighborhood ($7\leq R_{\rm gal}<9\kpc$, $0\leq|z|<0.5\kpc$). The black points with error bars indicate the mode of the abundance data in 1 Gyr-wide age bins, and the gray error bars along the bottom of each panel indicate the median age and abundance errors as a function of age.}
    \label{fig:yield-outflow}
\end{figure}

@@

We further highlight the discrepant age patterns in the bottom panels of Figure \ref{fig:lmr-ages}, which compare the overall stellar age distribution against that of the locally metal rich (LMR) stars, defined here as $\mathFeH\ge+0.1$.\footnote{
    While the amplitude of the peaks is sensitive to the precise location of the LMR cutoff, the dearth of intermediate-age stars is robust to adjustments of $\pm0.05\dex$.
} For APOGEE, the distributions are similar, both peaking near $\sim5\Gyr$, although very few of the LMR stars have ages $\gtrsim10\Gyr$. Our two-infall models predict an overall age distribution that is similar to the data, but both models predict a distinctly bimodal age distribution for LMR stars. The two populations reflect the metal-rich endpoints of the two successive infall epochs. The trough between the modes lies at $\sim5\Gyr$ for both models, right where the APOGEE distribution peaks. It is difficult to predict a unimodal age distribution with the available parameter space. A dearth of intermediate-age, high-metallicity stars is a natural outcome of a substantial dilution event, which is a defining feature of the two-infall scenario. None of the adjustments that we explored in Figure \ref{fig:abundance-evolution-params} substantially increase the proportion of intermediate-age stars.

@@

Radial gas flows are a potential alternative to the outflow prescription that we use in this paper. Within the two-infall paradigm, \citet{spitoni_effects_2011} found that inward flow velocities of a $\sim$ few $\kms$ can steepen the predicted radial metallicity gradient to match observations. Similarly, \citet{palla_chemical_2020,palla_mapping_2024} used an inward flow with a speed of $\sim1\kms$ to reproduce the observed radial metallicity profile in the absence of Galactic outflows. Johnson et al. (2025, in preparation) considered several possible scenarios for the processes that drive radial gas flows. Their findings support these previous arguments that radial gas flows should regulate the overall metal abundance and radial gradient, even beyond the two-infall scenario.

In our models, the outflows regulate the overall metallicity in a radially dependent manner, which \citet{johnson_milky_2024} demonstrated to improve agreement with age trends across the Galactic disk. Johnson et al. (2025, in preparation) found similar results for radial gas flows with velocities that are relatively constant in both Galactocentric radius and time. At least one radial gas flow scenario therefore has similar effects on metal enrichment as our outflow prescription. However, the outflow is simpler in implementation; we can simply adjust the mass loading factors up or down in combination with a given scale of yields. This approach allows us to easily consider models that predict metallicity to evolve on different timescales but reach similar abundances at the present day (see, e.g., Figure \ref{fig:yield-outflow}). A radial gas flow requires at least one assumption about the processes that drive the flow and generally do not afford the same level of flexibility. We have therefore focused on models with outflows in this paper for the sake of simplicity, and we clarify that some, but not all, radial gas flow models should have similar effects.
